---
format: html
---

```{r}
#| echo: false
sec = 2
cur = 0
```

# Extreme value theory

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` Maximum values

Let $X_1,\ldots,X_n$ be independent identically distributed random variables (i.i.d. r.v.) with the cumulative distribution function $F(x)=F_X(x)$. Denote
$$
X_M=X_{M,n}= \max\{X_1,\ldots,X_n\}.
$$ {#eq-2.1}
Then
$$
\begin{aligned}
\mathbb{P}(X_M\leq x) &= \mathbb{P}(X_1\leq x, \ldots, X_n\leq x) \\
&= \mathbb{P}(X_1\leq x) \ldots \mathbb{P}(X_n\leq x) \\
&= \bigl(\mathbb{P}(X\leq x)\bigr)^n = \bigl( F(x) \bigr)^n.
\end{aligned}
$$ {#eq-2.2}
We can attempt to standardise the values of $X_M$: consider, for some $\alpha_n>0$ and $\beta_n\in\mathbb{R}$,
$$
\begin{aligned}
\mathbb{P}\biggl(\frac{X_M-\alpha_n}{\beta_n}\leq x\biggr) &=\mathbb{P}(X_M\leq \beta_n x + \alpha_n)\\
& = \bigl(F(\beta_n x + \alpha_n)\bigr)^n.
\end{aligned}
$$ {#eq-2.3}
:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-important icon=false}

## `{r} sec`.`{r} cur` Theorem: Extreme value theorem (Fisher--Tippett--Gnedenko theorem)

If there exists $\alpha_n>0$ and $\beta_n$, such that there exists
$$
G(x):= \lim_{n\to\infty}\bigl(F(\beta_n x + \alpha_n)\bigr)^n
$$ {#eq-2.4}
such that $G(x)$ is a non-degenerate cumulative distribution function (i.e. $G(x)$ is **not** simultaneously identically equal to $0$ for all $x<x_1$ and identically equal to $1$ for all $x>x_1$ with some $x_1$), then this limiting function $G(x)$ may have **only one of the two following forms** :

- either, for some $\gamma\neq0$, $a,b\in\mathbb{R}$
$$
G(x):=G_\gamma(x) := \exp\Biggl(-\biggl(1+\gamma \frac{x-a}{b} \biggr)^{-\frac{1}{\gamma}}\Biggr),
$$ {#eq-2.5}

- or
$$
G(x):=G_0(x):=\exp\Biggl(-\exp\biggl(-\frac{x-a}{b} \biggr)\Biggr).
$$ {#eq-2.6}

Note that, for $\gamma<0$, $G_\gamma(x)$ corresponds to a Weibull distribution. For $\gamma>0$, the corresponding distribution is called Fr\'echet distribution, and for $\gamma=0$, it's called Gumbel distribution.

+--------+-------------------------+--------------+-------------------------+
| $G(x)$ | Weibull \               | Gumbel \     | Fr'echet \              |
|        | $\gamma<0$              | $\gamma=0$   | $\gamma>0$              |
+--------+-------------------------+--------------+-------------------------+
| $F(x)$ | Uniform \               | Exponential\ | Burr\                   |
|        | Beta                    | Gamma\       | Pareto\                 |
|        |                         | Log-normal\  | $t$\                    |
|        |                         | Normal\      | Log-gamma               |
|        |                         | Weibull      |                         |
+--------+-------------------------+--------------+-------------------------+
| $x$    | $x<a-\dfrac{b}{\gamma}$ | $\mathbb{R}$ | $x>a-\dfrac{b}{\gamma}$ |
+--------+-------------------------+--------------+-------------------------+

:::