---
format: html
---

```{r}
#| echo: false
sec = 4
cur = 0
```

# Reinsurance

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` Notations

- $X$ is the gross claim amount random variable

- $Y$ is the net claim amount (paid by the main insurer after receiving the reinsurance recovery)

- $Z$ is the amount paid by the reinsurer

:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-warning icon=false}

## `{r} sec`.`{r} cur` Example: Excess of loss reinsurance

The insurer pays in full up to an amount $M$, the reinsurer pays all above $M$, if needed:
$$
Y=\begin{cases}
X, & \text{if } X\leq M,\\
M, & \text{if } X>M;
\end{cases} \qquad Z=X-Y.
$$
Then
$$
\begin{aligned}
m_{k,Y} &= \int_0^M x^k f_X(x)\,dx + M^k \mathbb{P}(X>M),\\
M_Y(t)&= \int_0^M e^{tx}f_X(x)\,dx +e^{tM}\mathbb{P}(X>M),
\end{aligned}
$$
and
$$
\begin{aligned}
m_{k,Z} &= \int_M^\infty (x-M)^k f_X(x)\,dx, \\
M_Z(t) &= \mathbb{P}(X\leq M) +\int_M^\infty e^{t(x-M)}f_X(x)\,dx.
\end{aligned}
$$
:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` Reinsurer's view

In the conditions of the previous example, the reinsurer pays $Z$, evidently, only if $Z=X-Y>0$, otherwise, the reinsurer even does not know that a claim happened.

Hence, the distribution function for reinsurer is
$$
\begin{aligned}
F_Z(z)&=\mathbb{P}(Z\leq z\mid Z>0)=\mathbb{P}(X\leq z+M\mid X>M)\\
&= \frac{\mathbb{P}(M<X\leq z+M)}{\mathbb{P}(X>M)}= \frac{F_X(z+M)-F_X(M)}{1-F_X(M)},
\end{aligned}
$$
that looks familiar (see formula (1.1)).
Then
$$
f_Z(z)= \frac{f_X(z+M)}{1-F_X(M)}, \quad z>0.
$$
:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-warning icon=false}

## `{r} sec`.`{r} cur` Example: Proportional reinsurance

For $0<\alpha<1$, we set
$$
Y=\alpha X, \qquad Z=(1-\alpha)X.
$$
Then e.g.
$$
\begin{alignat*}{2}
\mathbb{E}(Y)&=\alpha\mathbb{E}(X), &\qquad
\mathbb{E}(Z)&=(1-\alpha)\mathbb{E}(X),\\
\mathrm{Var}(Y)&=\alpha^2\mathrm{Var}(X), &\qquad
\mathrm{Var}(Z)&=(1-\alpha)^2\mathrm{Var}(X).
\end{alignat*}
$$
:::

