---
format: html
---

```{r}
#| echo: false
sec = 6
cur = 0
```

# Survival models

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` Main objects of study and notations

- The **future lifetime of a newborn person** is a  *continuous*  random variable $T:\Omega\to[0,\mathfrak{w} ]$, where $\mathfrak{w} $ is the **limiting age**.

- Distribution function of $T$ is denoted
$$
F(t):=\mathbb{P}(T\leq t), \qquad t\geq0.
$$

- **Survival function** of $T$ is
$$
S(t):=\mathbb{P}(T>t)=1-F(t), \qquad t\geq0.
$$

- The **future lifetime after age** $x$, for $x\in[0,\mathfrak{w} ]$, is a random variable $T_x:\Omega\to [0,\mathfrak{w} -x]$ (i.e. $T=x+T_x$). In particular, $T_0=T$.

- Distribution function of $T_x$ is denoted
$$
F_x(t) :=\mathbb{P}(T_x\leq t).
$$

- Survival function of $T_x$ is denoted
$$
S_x(t) :=\mathbb{P}(T_x> t)=1-F_x(t).
$$

- Relation between $F_x$ and $F$ (or $S$):
$$
\begin{aligned}
F_x(t)&=\mathbb{P}(T\leq x+t\mid T>x)=\frac{\mathbb{P}(x<T\leq x+t)}{\mathbb{P}(T>x)}\\
& =\frac{F(x+t)-F(x)}{S(x)}=\frac{F(x+t)-F(x)}{1-F(x)}.
\end{aligned}
$$ {#eq-Fxt}

:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` Actuarial notations

- ${}_{t}q_{x}:= F_x(t)$

- ${}_{t}p_{x}:=1-{}_{t}q_{x}=S_x(t)$

- $q_{x}:={}_{1}q_{x}$

- $p_{x}:={}_{1}p_{x}$

:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` Survival probabilities relations

- We have
$$
\begin{aligned}
S_x(t) &= \mathbb{P}(T_x>t) = \mathbb{P}(T>x+t\mid T>x)\\
&=\frac{\mathbb{P}(T>x+t)}{\mathbb{P}(T>x)}=\frac{S(x+t)}{S(x)}.
\end{aligned}
$$ {#eq-eqkqwcsae1}

- In actuarial notations,
$$
{}_{t}p_{x}=\frac{{}_{x+t}p_{0}}{{}_{x}p_{0}}.
$$

- Therefore,
$$
{}_{s+t}p_{x}=\frac{{}_{x+s+t}p_{0}}{{}_{x}p_{0}}=\frac{{}_{x+s}p_{0}}{{}_{x}p_{0}}\cdot
\frac{{}_{x+s+t}p_{0}}{{}_{x+s}p_{0}}={}_{s}p_{x}\cdot {}_{t}p_{x+s}.
$$
As a result,
$$
{}_{s+t}p_{x}={}_{s}p_{x}\cdot {}_{t}p_{x+s}
={}_{t}p_{x}\cdot {}_{s}p_{x+t}.
$$ {#eq-consist}

- This (obvious) property is called the **consistency condition**.

:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` The force of mortality

- The **force of mortality** $\mu_x$ represents the instantaneous rate of mortality at a certain age $x\in[0,\mathfrak{w} ]$:
$$
\mu_x = \lim_{h\to0} \frac1h \mathbb{P}(T\leq x+h\mid T>x).
$$

- By ([-@eq-Fxt]), $F_x(0)=0$ and
$$
\begin{aligned}
\mu_x &= \lim_{h\to0} \frac1h \bigl( F_x(h)-F_x(0) \bigr)\\
&=F_x'(0)=\dfrac{d}{dt} \bigl( {}_{t}q_{x} \bigr)\biggr\rvert_{t=0}.
\end{aligned}
$$

- Consider the probability density function of $T$:
$$
f(t):=F'(t).
$$
Then, one gets from ([-@eq-Fxt]):
$$
\mu_x = \frac{F'(x)}{1-F(x)}= \frac{f(x)}{1-F(x)}.
$$

- The term "force of mortality" is mainly used in actuarial science. In engineering, the term **failure rate** is in use, and in statistics, it is called the **hazard function**.

- Since $S(x)=1-F(x)$, one can rewrite further:
$$
\mu_x=-\, \frac{S'(x)}{S(x)}=-\, \frac{d}{dx} \bigl( \ln S(x) \bigr).
$$ {#eq-eqkqwcsae}

:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-tip icon=false}

## `{r} sec`.`{r} cur` Remark

The following interpretation may be useful: since $F_x(h)={}_{h}q_{x}$, we have
$$
{}_{h}q_{x}\approx \mu_x \cdot h.
$$
:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` Probability density function of $T_x$

- We consider
$$
f_x(t)=\dfrac{d}{dt} F_x(t).
$$

- We have then, by ([-@eq-Fxt]),
$$
\begin{aligned}
f_x(t)& = \dfrac{d}{dt} \frac{F(x+t)-F(x)}{S(x)}= \frac{F'(x+t)}{S(x)}\\
& = \frac{F'(x+t)}{S(x+t)}\frac{S(x+t)}{S(x)}
\end{aligned}
$$
and since $F'(y)=1-S'(y)$, we can use ([-@eq-eqkqwcsae]) and ([-@eq-eqkqwcsae1]) to continue:
$$
\begin{aligned}
&=-\frac{S'(x+t)}{S(x+t)}S_x(t)\\& =S_x(t)\mu_{x+t},
\end{aligned}
$$
i.e., for $0\leq t\leq \mathfrak{w} -x$,
$$
f_x(t)=S_x(t)\mu_{x+t}={}_{t}p_{x} \cdot \mu_{x+t}.
$$ {#eq-densrepr}

From ([-@eq-densrepr]), we have
$$
\begin{aligned}
\int_0^t {}_{s}p_{x} \cdot \mu_{x+s}\,ds&=
\int_0^t f_x(s)\,ds = \int_0^t \dfrac{d}{dt} F_x(s)\,ds\\
& = F_x(t)-F_x(0) =F_x(t);
\end{aligned}
$$
i.e.
$$
{}_{t}q_{x}= \int_0^t {}_{s}p_{x} \cdot \mu_{x+s}\,ds.
$$ {#eq-qxexp}
:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` Integral expressions

We found in ([-@eq-qxexp]) an integral expression for ${}_{t}q_{x}$.
Consider now another one, for ${}_{t}p_{x}$.
By ([-@eq-eqkqwcsae]),
$$
\mu_{x+s} = -\frac{S'(x+s)}{S(x+s)} =-\frac{d}{ds} \ln S(x+s),
$$
hence, integrating in $s\in[0,t]$ and using ([-@eq-eqkqwcsae1]),
$$
\begin{aligned}
-\int_0^t \mu_{x+s}\,ds &= \Bigl[\ln S(x+s)\Bigr]_0^t=\ln \frac{S(x+t)}{S(x)}\\
& = \ln S_x(t)=\ln{}_{t}p_{x}.
\end{aligned}
$$
Therefore,
$$
{}_{t}p_{x} = \exp\biggl(-\int_0^t \mu_{x+s}\,ds \biggr).
$$ {#eq-tpxInt}
:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-tip icon=false}

## `{r} sec`.`{r} cur` Remark

Formula ([-@eq-tpxInt]) can be also obtained by considering a time-non-homogeneous MJP with states $1$, \textit{Alive}, and $2$,  *Dead*, and the transition rate $\mu_t$. Then
$$
{}_{t}p_{x}=p_{\overline{1,1}}(x,x+t)=\exp\biggl(-\int_{x}^{x+t}\mu_s\,ds\biggr),
$$
that is nothing but ([-@eq-tpxInt]), by a change of variables.
:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` Life table functions

- $l_x$ denotes the expected number of lives at (integer) age $x\in[0,\mathfrak{w}]$.

- $d_x$ is the expected number of deaths between the ages of $x$ and $x+1$.

- The following formulas holds:
$$
\begin{aligned}
d_x&= l_x-l_{x+1};\\
p_x&=\dfrac{l_{x+1}}{l_x}; \\
q_x&=1-p_x=1-\dfrac{l_{x+1}}{l_x}=\dfrac{d_x}{l_x};\\
{}_{t}p_{x}&= \dfrac{l_{x+t}}{l_x}; \\
{}_{t}q_{x}& = 1-{}_{t}p_{x}=\dfrac{l_x-l_{x+t}}{l_x}.
\end{aligned}
$$ {#eq-px-lx}

:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` Justification of formula ([-@eq-px-lx])

Let $Y_x$ be the random variable of the alive individuals of age $x$, so that
$$
l_x=\mathbb{E}(Y_x), \quad x\in \mathbb{Z}_+.
$$
Then, by the property of the conditional expectation
$$
l_{x+1}=\mathbb{E}(Y_{x+1})=\mathbb{E}\bigl( \mathbb{E}(Y_{x+1} \mid Y_x=l_x) \bigr).
$$
Next, $Z_x:=\mathbb{E}(Y_{x+1} \mid Y_x=l_x)$ is a random variable which satisfies the binomial law: there are $l_x$ individuals of age $x$ each of them may survive in the next year (independently from others) with probability $p_x$ and die with probability $q_x=1-p_x$, i.e. e.g.
$$
\mathbb{P}(Z_x = k) = \binom{l_x}{k} (p_x)^k (q_x)^{l_x-k}.
$$
We know from Probability course that the expectation of a binomial random variable is the product of the number of possible trials and the probability of "success", therefore,
$$
l_{x+1}=\mathbb{E}(Y_{x+1}) = \mathbb{E} (Z_x) = l_x p_x,
$$
that implies ([-@eq-px-lx]).
:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-important icon=false}

## `{r} sec`.`{r} cur` Theorem: Uniform distribution of deaths

If deaths are uniformly distributed between the ages of $x$ and $x+1$, then, **for all** $t\in[0,1]$:
$$
{}_{t}q_{x}=t \cdot q_{x}.
$$

::: {.proof}
Since the number of deaths is uniformly distributed, $l_{x+t}\in[l_{x+1},l_x]$ for all $t\in[0,1]$ and
$l_{x+t}= (1-t)l_{x}+t l_{x+1}$,

therefore,
$$
{}_{t}q_{x}=\dfrac{l_x-l_{x+t}}{l_x}=\dfrac{t l_x - tl_{x+1}}{l_x}
= t(1-p_x)=tq_x,
$$
that proves the statement.
:::

::::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-tip icon=false}

## `{r} sec`.`{r} cur` Remark:  Constant force of mortality assumption

Another assumption for the ages between $x$ and $x+1$ is that the force of mortality remains constant:
$$
\mu_{x+s} = \mu_x, \quad s\in[0,1).
$$
In this case, by ([-@eq-tpxInt]), one has
$$
{}_{t}p_{x} = e^{-t\mu_x} =
\bigl( e^{-1\cdot \mu_x}\bigr)^t
=(p_x)^t.
$$
Therefore,
$$
{}_{t}q_{x}=1 - e^{-t\mu_x} \neq t
(1-e^{-\mu_x})= t q_x.
$$
However, if $\mu_x\approx 0$, then
${}_{t}q_{x}\approx t\mu_x \approx t q_x$.
:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` Analogue of consistency condition

The consistency condition ([-@eq-consist]) is actually nothing but the conditional probability identity. It's counterpart for ${}_{t}q_{x}$ reads as follows:
$$
{}_{t+s}q_{x}= {}_{s}q_{x}+{}_{s}p_{x}\cdot{}_{t}q_{x+s}.
$$ {#eq-consq}
The interpretations is as follows: to die within the next $t+s$ years the person may either die in the next $s$ years or survive in the next $s$ years but die within next $t$ years thereafter.

Formula ([-@eq-consq]) can be derived rigorously from the law of the full probability (do this by yourself!) or just from ([-@eq-Fxt]):
$$
\begin{aligned}
{}_{t+s}q_{x}&=
\frac{F(x+t+s)-F(x)}{1-F(x)}\\
&= \frac{F(x+t+s)-F(x+s)}{1-F(x)}
+\frac{F(x+s)-F(x)}{1-F(x)}\\
&= \frac{F(x+t+s)-F(x+s)}{1-F(x+s)}\frac{1-F(x+s)}{1-F(x)}
\\&\quad +\frac{F(x+s)-F(x)}{1-F(x)}\\
& = {}_{t}q_{x+s}\cdot{}_{x}p_{s} + {}_{x}q_{s},
\end{aligned}
$$
where we used that, by ([-@eq-eqkqwcsae1]),
$$
\frac{1-F(x+s)}{1-F(x)} = \frac{S(x+s)}{S(x)}={}_{s}p_{x}.
$$
:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` Central rate of mortality

The **central rate of mortality** at age $x$ is the ration of the number of deaths over the year of age $x$ to $x+1$ to the average number of lives then:
$$
m_x=\dfrac{d_x}{\displaystyle\int_0^1 l_{x+t}\,dt}.
$$
We can also rewrite this:
$$
m_x= \frac{q_x}{\displaystyle\int_0^1 {}_{t}p_{x} \,dt}=
\dfrac{\displaystyle\int_0^1 {}_{t}p_{x} \mu_{x+t}\,dt}{\displaystyle\int_0^1 {}_{t}p_{x} \,dt}<q_x.{}
$$
:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` Complete expectation of life

The **expected future lifetime** after age $x\in[0,\mathfrak{w}]$ is defined as follows:
$$
\mathring{e}_x: = \mathbb{E}(T_x).
$$
Using the definition of the expectation for a continuous random variable, we have, by ([-@eq-densrepr]),
$$
\begin{aligned}
\mathbb{E}(T_x)&= \int_0^{\mathfrak{w}-x} t f_x(t)\,dt\\
& = \int_0^{\mathfrak{w}-x} t \cdot \dfrac{d}{dt} F_x(t)\,dt
\\
&= -\int_0^{\mathfrak{w}-x} t \cdot \dfrac{d}{dt} {}_{t}p_{x} \,dt
\end{aligned}
$$
and then we integrate by parts
$$
\begin{aligned}
&=- \Bigl[t {}_{t}p_{x} \Bigr]_0^{\mathfrak{w}-x}+
\int_0^{\mathfrak{w}-x} {}_{t}p_{x} \,dt\\
&= \int_0^{\mathfrak{w}-x} {}_{t}p_{x} \,dt,
\end{aligned}
$$
as $ {}_{\mathfrak{w}-x}p_{x}=\mathbb{P}(T_{x}>\mathfrak{w}-x)=0$.

Since
$$
{}_{t}p_{x}=0 \quad\text{for } t>\mathfrak{w}-x
$$
as well, one can rewrite:
$$
\mathring{e}_x = \mathbb{E}(T_x)=\int_0^\infty {}_{t}p_{x} \,dt.
$$
:::

```{r}
#| echo: false
cur = cur + 1
```

:::: {.callout-important icon=false}

## `{r} sec`.`{r} cur` Theorem: Monotonicity of the lifetime

Function $x+\mathring{e}_x$ is increasing in $x$, i.e.
$$
x+\mathring{e}_x<y+\mathring{e}_y, \quad x<y.
$$
In other words, expectation for the age of death $T=x+T_x$ increases when the person managed to live more.

::: {.proof}
We have
$$
\begin{aligned}
\frac{d}{dx}\mathring{e}_x &= \int_0^\infty \frac{d}{dx} {}_{t}p_{x}\,dt=
\int_0^\infty \frac{d}{dx} \frac{S(x+t)}{S(x)}\,dt\\
& = \int_0^\infty \frac{S'(x+t)S(x)-S'(x)S(x+t)}{(S(x))^2}\,dt\\
& =\int_0^\infty \frac{S'(x+t)}{S(x)}\,dt
+\mu_x\int_0^\infty \frac{S(x+t)}{S(x)}\,dt \\
& = -1+\mu_x \mathring{e}_x.
\end{aligned}
$$
Therefore,
$$
\frac{d}{dx} \left( \mathring{e}_x + x \right) = \mu_x \mathring{e}_x>0.
$$
:::
::::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-tip icon=false}

## `{r} sec`.`{r} cur` Remark

The previous statement may be confusing if we look at equalities
$T=x+T_x$ and $T=y+T_y$,
so then we would expect that $
\mathbb{E}(T)=x+\mathring{e}_x$ and $\mathbb{E}(T)=y+\mathring{e}_y$. Apparently, these are two different random variables both (actually, wrongly) denoted by $T$. Namely, when we write $T=x+T_x$, we consider the random lifetime $T$ conditional that a person was alive at age $x$, it is measurable w.r.t. $\sigma$-algebra generated by ages $\leq x$. The random variable $\tilde{T}=y+T_y$ is, hence, different.
:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` Curtate expectation of life

The **curtate future lifetime** of a life age $x$ is
$$
K_x=[T_x],
$$
where $[y]$ denotes the integer part of $y$, i.e. $y$ is rounded down to the nearest integer.

We have then, because of the continuity of $T$ (i.e. of $F(t)$),
$$
\begin{aligned}
\mathbb{P}(K_x=k)& =\mathbb{P}(k\leq T_x<k+1)\\
&=\mathbb{P}(k<T_x\leq k+1)\\&={}_{k}p_{x}\cdot q_{x+k}.
\end{aligned}
$$
We define the **curtate expectation of life** as follows:
$$
e_x=\mathbb{E}(K_x).
$$
We have, hence,
$$
\begin{aligned}
e_x &= \sum_{k=0}^{[\mathfrak{w}-x]} k\cdot \mathbb{P}(K_x=k)
=\sum_{k=1}^{[\mathfrak{w}-x]} \sum_{j=1}^k 1\cdot \mathbb{P}(K_x=k)\\
&=\sum_{j=1}^{[\mathfrak{w}-x]}\sum_{k=j}^{[\mathfrak{w}-x]}\mathbb{P}(K_x=k)= \sum_{j=1}^{[\mathfrak{w}-x]}\mathbb{P}(K_x\geq j).
\end{aligned}
$$
Since $\mathbb{P}(K_x\geq j)= \mathbb{P}(T_x\geq j)={}_{j}p_{x}$ ($=0$ for $j>[\mathfrak{w}-x]$),
$$
e_x =\mathbb{E}(K_x) =\sum_{j=1}^{\infty}{}_{j}p_{x}.
$$
:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` Simple parametric survival models

- **Exponential model** :
$$
\mu_x=\mu, \quad \text{for all } x\in[0,\mathfrak{w}].
$$
Then
$$
{}_{t}p_{x}=S_x(t)=e^{-\mu t}
$$
Note that
$$
f(t) = S(t)\mu_t = \mu e^{-\mu t}.
$$

- **Weibull model** :
$$
\mu_x=\alpha \beta x^{\beta-1}, \quad \alpha, \beta>0.
$$
Note that $\mu_x$ is increasing if $\beta>1$ and decreasing if $\beta\in(0,1)$. $\beta=1$ corresponds to the exponential model.
We have
$$
{}_{t}p_{x}=S_x(t)=\exp\Bigl( - \alpha \bigl((x+t)^\beta -x^\beta\bigr) \Bigr).
$$
In particular,
$$
S(t)=e^{- \alpha t^\beta},
$$
hence,
$$
f(t) = \alpha \beta t^{\beta-1}e^{- \alpha t^\beta}.
$$
Parameter $\beta$ is called **shape** and $\alpha$ is called **scale**.



- **Gompertz model** :
$$
\mu_x= b c^x, \quad b,c>0.
$$

If $c>1$, then $\mu_x$ is increasing; and if $c\in(0,1)$, then $\mu_x$ is decreasing. $c=1$ corresponds to the exponential model.

Parameter $b>0$ is called **rate**. Often the hazard rate is written for $c=e^\gamma$, where $\gamma=\ln c\in\mathbb{R}$ is called **shape**.

We have
$$
{}_{t}p_{x} = S_x(t)=g ^{c^x(c^t-1)},
$$
where
$$
g=\exp\Bigl(-\frac{b}{\ln c}\Bigr).
$$

We have also:
$$
f(t) = \mu_t S(t)= bc^t g^{c^t-1}.
$$

- **Gompertz--Makeham model** :
$$
\mu_x= a+b c^x, \quad a,b,c>0.
$$
Then
$$
{}_{t}p_{x} =S_x(t)=e^{-at} g ^{c^x(c^t-1)},
$$
where $g$ is the same as above.
Then
$$
f(t) = (a+b c^t) e^{-at} g ^{c^t-1}.
$$

:::

