---
format: html
---

```{r}
#| echo: false
sec = 5
cur = 0
```

# Compound distributions

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` The collective risk model

The number $N$ of claims (raised during an interval of time) is, usually, random, i.e. we may treat it as a random variable
$$
N:\Omega\to\mathbb{Z}_+:=\mathbb{N}\cup\{0\}.
$$

Assume that all claims, $X_1$, \ldots, $X_N$ are i.i.d.r.v. (independent identically distributed random variables).

The total sum of all claims is then
$$
S=S_N:=X_1+\ldots+X_N,
$$
where $S:=0$ if $N=0$.

Then
$$
F_S(x)=\sum_{n=0}^\infty \mathbb{P}(N=n)\, \mathbb{P}(S\leq x \mid N=n).
$$

Next,
$$
\begin{aligned}
\mathbb{E}(S) &= \mathbb{E}(N)\, \mathbb{E}(X);\\
\mathrm{Var}(S) &= \mathbb{E}(N)\, \mathrm{Var}(X) + \mathrm{Var}(N)\, \left( \mathbb{E}(X) \right)^2;\\
M_S(t)& = M_N\left( \ln M_X(t) \right).
\end{aligned}
$$ {#eq-eq1f}
:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` The Poisson distribution (reminder)

Recall that $N:\Omega\to\mathbb{Z}_+$ has the **Poisson distribution** with parameter $\lambda>0$, notation: $N\sim Po(\lambda)$, if
$$
\mathbb{P}(N=n)=\frac{\lambda^n}{n!}e^{-\lambda}, \qquad n\in\mathbb{Z}_+.
$$

Then
$$
\begin{aligned}
\mathbb{E}(N)&=\mathrm{Var}(N)=\lambda,\\
M_N(t)&= \exp\bigl( \lambda(e^t-1) \bigr).
\end{aligned}
$$ {#eq-mgfpo}
:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` The compound Poisson distribution

Let $N\sim Po(\lambda)$, let $X_1, \ldots, X_N$ are i.i.d.r.v., let $m_k:=m_{k,X}$.
Then $S=X_1+\ldots+X_N$ has the **compound Poisson distribution** with the parameter $\lambda$ and
$$
\begin{aligned}
\mathbb{E}(S)&=\lambda m_1,\\
\mathrm{Var}(S)&=\lambda m_2,\\
M_S(t)&= \exp\bigl( \lambda (M_X(t)-1) \bigr).
\end{aligned}
$$ {#eq-mgfcpo}
:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` Sums of independent compound Poisson r.v.

Let $S_1,\ldots, S_m$ be independent compound Poisson random variables with parameters $\lambda_i$ and CDF of the single claims $F_i(x)$, $1\leq i\leq m$, i.e.
$$
S_i = X_1^{(i)}+\ldots+X_{N_i}^{(i)},
$$
where, for each $i$, $\{X_j^{(i)}\}$ are i.i.d.r.v. with CDF $F_i$, and $N_i\sim Po(\lambda_i)$. Moreover, $\{X_j^{(i)}\}$ are also independent for different $i$.
Then
$$
\mathbf{S}:=S_1+\ldots+S_m
$$
is a compound Poisson random variable with the parameter
$$
\Lambda=\lambda_1+\ldots+\lambda_m
$$
and the CDF of the single claim
$$
F(x)=\frac{1}{\Lambda} \bigl( \lambda_1 F_1(x)+\ldots+\lambda_m F_m(x) \bigr),
$$
i.e.
$$
S=Y_1+\ldots+Y_N
$$
for i.i.d.r.v. $\{Y_j\}$ with the CDF $F(x)$, and $N\sim Po(\Lambda)$.
:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` The compound binomial distribution

$N\sim Bin(n,p)$, $n\in\mathbb{N}$, $p\in[0,1]$, iff, for $0\leq k\leq n$,
$$
\mathbb{P}(N=k)=\binom{n}{k} p^k(1-p)^{n-k}.
$$
Then
$$
\begin{gathered}
\mathbb{E}(N) = np, \qquad \mathrm{Var}(N)=np(1-p), \\ M_N(t)=(pe^t+1-p)^n.
\end{gathered}
$$
We can define $S=X_1+\ldots+X_N$ for i.i.d.r.v. $X_i$, and the distribution of $S$ is called the **compound binomial distribution**. Then
$$
\begin{gathered}
\mathbb{E}(S)=np m_{1,X}, \quad \mathrm{Var}(S)= np m_{2,X} -np^2 m_{1,X}^2,\\
M_S(t) = (pM_X(t)+1-p)^n.
\end{gathered}
$$
:::

```{r}
#| echo: false
cur = cur + 1
```

::: {.callout-note icon=false}

## `{r} sec`.`{r} cur` Agrregated claims with reinsurance

- The aggregated claim $S=X_1+\ldots+X_N$ of a collective risk model can be done when the reinsurance is in force.

- Consider the case when the reinsurance is applied to each individual claim.

- Let $Y_i$ be the amount paid by the main insurer for the claim $X_i$, and $Z_i$ be the claim paid by the reinsurer, so $X_i=Y_i+Z_i$.

- Let $S_I=Y_1+\ldots+Y_N$ be the aggregated claim amount paid by the insurer, and $S_R=Z_1+\ldots+Z_N$ be the aggregated claim amount paid by the reinsurer.

- Of course, formulas ([-@eq-eq1f]) are still true if we replace there $S,X$ by $S_I,Y$ or by $S_R,Z$, respectively.

- Note that, for the **individual excess of loss reinsurance**, the reinsurer pays $Z_i$ only if $Z_i=X_i-M>0$. We can introduce $J_i=1$ if $X_i>M$ and $J_i=0$ otherwise. Then
$$
N_R:=J_1+\ldots+J_N
$$
is the number of claims paid by the reinsurer.

- $X_1,\ldots,X_N$ are i.i.d.r.v., let
$$
\mathbb{P}(J_i=1)=\mathbb{P}(X_i>M)=:\rho.
$$
Then (actually $J_i\sim Bin(1,\rho)$)
$$
M_{J_i}(t)=\mathbb{E}(e^{t J_i})= \rho e^t +(1-p)e^0 = \rho e^t +1-p,
$$
and for the compound random variable $N_R$,
$$
M_{N_R}(t)=M_N\bigl(\ln (\rho e^t +1-p)\bigr).
$$ {#eq-NRdistr}

- Let $\{i_1,\ldots,i_{N_R}\}\subset\{1,\ldots, N\}$ be indexes such that $X_{i_k}>M$ for $1\leq k\leq N_R$, and set $W_k:=X_{i_k}-M$. Then there is another representation
$$
S_R = W_1+\ldots + W_{N_R}.
$$

- If, additionally, $N\sim Po(\lambda)$ for a $\lambda>0$, then ([-@eq-NRdistr]) reads, by ([-@eq-mgfpo]) or ([-@eq-mgfcpo]),
$$
M_{N_R}(t) =\exp\Bigl(\lambda \bigl(M_{J_i}(t)-1\bigr)\Bigr) =\exp\bigl(\rho\lambda (e^t-1)\bigr),
$$
i.e. $N_R\sim Po(\rho\lambda)$, by ([-@eq-mgfpo]). (That is nothing but the **thinning property** of the Poisson distribution.)

:::

